Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-05-17 manuscript/revision2/revision2.tex
revision2: Add new SI Datasets 1-6 to the SI Appendix

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, Gulf vegetative growth
occurred before Midwestern vegetative growth, and Gulf flowering
occurred after Midwestern flowering (Figure~\ref{fig-map} A). At the
four northernmost common gardens (hereafter `North' gardens), located
within the natural range of the Midwest subpopulation, both Gulf
vegetative growth and flowering occurred after Midwest vegetative growth
and flowering. At the Oklahoma common garden, located near the natural
range limits of both the Gulf and the Midwest subpopulations, Gulf and
Midwest vegetative growth occurred over the same period, and Gulf
flowering occurred after Midwestern flowering (Figure~\ref{fig-map} A).
These patterns led to strong negative phenotypic correlations for the
timing of the start of vegetative growth between the North and Texas
gardens, particularly in the Gulf and across all individuals from the
Gulf and Midwest subpopulations (hereafter, `Both' subpopulations), and
contributed to positive phenotypic correlations for the timing of
flowering that had larger magnitudes at more northern gardens
(Figure~\ref{fig-map} B).

@@

We used a greedy algorithm to select covariance matrices from each
model's set that significantly improved the model likelihood. Five
GxWeather covariance structures were selected, zero to two per
subpopulation: two for the start of vegetative growth, and three for
flowering date. Of these five, three had mass on them in mash models of
the strong effects (Figure~\ref{fig-covar} A-B). Two of these three
matrices had negative covariances between sites between Texas and North
gardens (Figure~\ref{fig-covar} A), while one had all positive or
near-zero covariances. SNP-associated phenotypic effects covaried with
different weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the mash model of Gulf vegetative growth fell on a
covariance matrix constructed using the covariance of day length 14 days
prior to the date of vegetative growth. The covariance matrix for this
weather cue was visually similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Mash
models of the timing of vegetative growth in the Midwest subpopulation
and Both subpopulations did not include this GxWeather covariance; the
Midwest had no weight on any GxWeather covariance, while Both
subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C) for vegetative growth.

@@

For greenup date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited antagonistic pleiotropy, or a difference in
effect sign, between pairs of Texas and North gardens
(Figure~\ref{fig-effects} A; SI Appendix, Fig SX). 78.7\% of pairwise
comparisons between North and Texas gardens had a difference in sign,
while only 28.6\% and 0.2\% of North-North or Texas-Texas comparisons
had a difference in sign, respectively (Figure~\ref{fig-effects}). The
majority of pairwise effects for greenup for the Midwest
(\textgreater55\%) and Both (\textgreater85\%) subpopulations were the
same sign, and effects most often differed in magnitude between gardens
within and between the regions (Figure~\ref{fig-effects} X).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited antagonistic pleiotropy, or a difference in effect
sign, within or between regions (Figure~\ref{fig-effects}). More effects
differed in magnitude between the Texas and North regions than within
these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects} X),
while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering, but a large proportion
of these differed in sign between Texas and North regions (42.7\%) or
within the North region (65.4\%). Finally, in Both subpopulations, less
than 20\% of pairwise effects differed in sign
(Figure~\ref{fig-effects}). Most differences in sign were between TX1,
the southernmost garden, and all other gardens (SI Appendix, Fig SX).
Similarly, more effects that differed in magnitude included gardens in
the Texas region (52.3-55.9\%), and most effect pairs in the North
region were not distinguishable (91.5\%).

@@

\caption{\label{fig-map}Characterization of green-up and flowering dates
from the switchgrass diversity panel. (a) Map and trait histograms of
green-up and flowering dates across two genetically distinct switchgrass
subpopulations and eight common gardens. Purple represents individuals
from the Midwest genetic subpopulation, and pink individuals from the
Gulf subpopulation; map positions represent the original collection
locations for the genotypes, and shapes represent the ecotype of the
genotype. Vertical dashed lines indicate the summer solstice. Common
gardens are arranged in latitudinal order. (b) Phenotypic correlations
between clonal replicates planted at eight common gardens, within and
between two genetic subpopulations. (c) Narrow sense heritability of
green-up and flowering within single common gardens (purple) and across
all eight common gardens (green), within and between two genetic
subpopulations. (d) Flow diagram of the methods applied to the green-up
and flowering dates to jointly estimate SNP effects across all sites.
Mash was fit to SNP effect data and used to find covariance matrices
that improved the mash model likelihood using a large set of randomly
selected, relatively unlinked SNP effects; this model was applied to a
``strong'' set of SNP effects with large effect sizes in the univariate
GWAS.}

@@

In 2019, we scored two phenological events every two days in two mapping
populations of switchgrass, a diversity panel and a pseudo-F2 cross,
planted at eight common garden locations (35, 39, 40). We scored the
start of vegetative growth as the day of the year when 50\% of the
tiller area of the crown of the plant cut the previous year had green
growth. The start of reproductive growth, or flowering date, was the day
of the year when 50\% of the plant tillers had panicles undergoing
anthesis. We scored green-up and flowering as day of the year, then
linked these dates to multiple weather-based environmental factors
measured daily at each common garden (SI Appendix, Section S1, Table
S1).

>>>>

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, Gulf vegetative growth
(hereafter `green-up') occurred before Midwestern green-up, and Gulf
reproductive growth (hereafter `flowering') occurred after Midwestern
flowering (Figure~\ref{fig-map} A). At the four northernmost common
gardens (hereafter `North' gardens), located within the natural range of
the Midwest subpopulation, both Gulf green-up and flowering occurred
after Midwest green-up and flowering. At the Oklahoma common garden,
located near the natural range limits of both the Gulf and the Midwest
subpopulations, Gulf and Midwest green-up occurred over the same period,
and Gulf flowering occurred after Midwestern flowering
(Figure~\ref{fig-map} A). These patterns led to strong negative
phenotypic correlations for the timing of the start of vegetative growth
between the North and Texas gardens, particularly in the Gulf and across
all individuals from the Gulf and Midwest subpopulations (hereafter,
`Both' subpopulations), and contributed to positive phenotypic
correlations for the timing of flowering that had larger magnitudes at
more northern gardens (Figure~\ref{fig-map} B).

@@

We used a greedy algorithm to select covariance matrices from each
model's set that significantly improved the model likelihood. Five
GxWeather covariance structures were selected, zero to two per
subpopulation: two for green-up date, and three for flowering date. Of
these five, three had mass on them in mash models of the strong effects
(Figure~\ref{fig-covar} A-B). Two of these three matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while one had all positive or near-zero
covariances. SNP-associated phenotypic effects covaried with different
weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the mash model of Gulf green-up fell on a
covariance matrix constructed using the covariance of day length 14 days
prior to the date of green-up. The covariance matrix for this weather
cue was visually similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Mash
models of the timing of vegetative growth in the Midwest subpopulation
and Both subpopulations did not include this GxWeather covariance; the
Midwest had no weight on any GxWeather covariance, while Both
subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C) for vegetative growth.

@@

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited antagonistic pleiotropy, or a difference in
effect sign, between pairs of Texas and North gardens
(Figure~\ref{fig-effects} B; SI Appendix, Fig SX). 78.7\% of pairwise
comparisons between North and Texas gardens had a difference in sign,
while only 28.6\% and 0.2\% of North-North or Texas-Texas comparisons
had a difference in sign, respectively (Figure~\ref{fig-effects} B). The
majority of pairwise effects for greenup for the Midwest
(\textgreater55\%) and Both (\textgreater85\%) subpopulations were the
same sign, and effects most often differed in magnitude between gardens
within and between the regions (Figure~\ref{fig-effects} B).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited antagonistic pleiotropy, or a difference in effect
sign, within or between regions (Figure~\ref{fig-effects} B). More
effects differed in magnitude between the Texas and North regions than
within these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects}
B), while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering, but a large proportion
of these differed in sign between Texas and North regions (42.7\%) or
within the North region (65.4\%). Finally, in Both subpopulations, less
than 20\% of pairwise effects differed in sign (Figure~\ref{fig-effects}
B). Most differences in sign were between TX1, the southernmost garden,
and all other gardens (SI Appendix, Fig SX). Similarly, more effects
that differed in magnitude included gardens in the Texas region
(52.3-55.9\%), and most effect pairs in the North region were not
distinguishable (91.5\%).

@@

\caption{\label{fig-map}Characterization of the timing of vegetative
(green-up) and reproductive (flowering) growth in the switchgrass
diversity panel. (a) Map and trait histograms of green-up and flowering
dates across two genetically distinct switchgrass subpopulations and
eight common gardens. Purple represents individuals from the Midwest
genetic subpopulation, and pink individuals from the Gulf subpopulation;
map positions represent the original collection locations for the
genotypes, and shapes represent the ecotype of the genotype. Vertical
dashed lines indicate the summer solstice. Common gardens are arranged
in latitudinal order. (b) Phenotypic correlations between clonal
replicates planted at eight common gardens, within and between two
genetic subpopulations. (c) Narrow sense heritability of green-up and
flowering within single common gardens (purple) and across all eight
common gardens (green), within and between two genetic subpopulations.
(d) Flow diagram of the methods applied to the green-up and flowering
dates to jointly estimate SNP effects across all sites. Mash was fit to
SNP effect data and used to find covariance matrices that improved the
mash model likelihood using a large set of randomly selected, relatively
unlinked SNP effects; this model was applied to a ``strong'' set of SNP
effects with large effect sizes in the univariate GWAS.}

@@

In 2019, we scored two phenological events every two days in two mapping
populations of switchgrass, a diversity panel and a pseudo-F2 cross,
planted at eight common garden locations (35, 39, 40). We scored the
start of vegetative growth, or green-up date, as the day of the year
when 50\% of the tiller area of the crown of the plant cut the previous
year had green growth. The start of reproductive growth, or flowering
date, was the day of the year when 50\% of the plant tillers had
panicles undergoing anthesis. We scored green-up and flowering as day of
the year, then linked these dates to multiple weather-based
environmental factors measured daily at each common garden (SI Appendix,
Section S1, Table S1).
